A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-11-06 latex/main.tex
Updates from Overleaf

The Delta, BA.1, and BA.2 lineages were 
particularly challenging, however, as the initial emergences of 
the saltation lineages were not densely sampled. In this case,
there is little information for the retrospective sampling grouping mechanism
to work with and it is very difficult to distinguish short-term time-travellers
from true outbreak samples. For these cases, we chose specific
high-quality samples with confident dates as ``seeds`` which are 
unconditionally included in the ARG on the specified date.

% We used more seeds, but in practise it doesn't matter and it's just
% complicating the story.
For Delta, we used two seeds:
one for B.1.617.1 (Kappa, which is sibling to Delta)
and one sample assigned to B.1.617.2.
[FIXME more Delta details and some references to back up our assertions.
There's some specific info in the Viridian paper on this,
I'm sure. 
We should mention long branch splitting here too.]
The resulting subgraph can be seen in Figure~\ref{fig:subgraph_delta}.

[FIXME: revise this paragraph to give some background info on Omicron, and reference
the Viridian work]
For Omicron BA.1 and BA.2,
we used SRR17041376 and SRR17461792, respectively.
We chose SRR17041376 (a sample dated 2021-11-12 from South Africa)
as the BA.1 seed (https://github.com/jeromekelleher/sc2ts-paper/issues/264),
as it could be well explained as as a non-recombinant
given the ARG available on its collection date.
For BA.2, we chose SRR17461792 (dated 2021-11-27 from South Africa),
which is the earliest BA.2 sample present in the Viridian v04 dataset.
It is dated later than SRR17041376 (the BA.1 seed),
allowing time for the BA.1 lineage to accumulate enough sampled descendants
and to get established in the ARG.
See Figure~\ref{fig:subgraph_omicron} for an overview of the resulting 
subgraph.

@@

Using the default ``parsimony improvement'' parameter of $p=3$, UShER+RIPPLES detected a total of 4111 recombination events.
Most of these subtend $<10$ samples (2616 singletons),
although four events subtended more than 100,000 samples.
Analysis of the descending samples for each event shows that
10 of the 44 Pango X lineages present in the dataset
% NOTE: listing them like this for sorting and manipulation purposes
(%
XA, 
XB, 
XAC, 
XAE, 
XAJ,
XBB, 
XBM, 
XC, 
XL, 
XW%
)
are associated unequivocally with a single recombination event.
7 lineages 
(%
XAM,
XBD, 
XBR, 
XE,
XM,
XS,
XY%
)
are associated with two events. Finally, 
one event subtending 1321 samples suggests that 
XAF,
XAL, 
XE, 
XH, 
XJ, 
XM, 
XY, 
and 
82 BA.2 samples share a recombination event [FIXME: With what breakpoint and parents?].
We also ran RIPPLES with $p=4$, which reduced the overall number of
events to 1174 (roughly 4-fold) but also approximately halved the number of
events involving X lineages, missing well-characterized events,
such as XBB (Table~\ref{tab:ripples_x_lineages}).

To compare these results with the sc2ts ARG, we first matched clades
by finding the most recent common ancestor (MRCA) of the samples
descending from each RIPPLES event. Overall, clade agreement was
high with [FIXME, get exact numbers] 37XX nodes (90\%) in the sc2ts ARG
where descendants match exactly. Of these nodes, 68\% had no recombinant
ancestry. Of the remainder, 595 could be matched with
one of the 855 recombination events in the sc2ts ARG
(Figure~\ref{fig:ripples_sc2ts_events}).
Table~\ref{tab:ripples_x_lineages} shows a summary of
all recombination events associated with Pango X lineages detected by RIPPLES,
and their relationship with the sc2ts ARG.
Of the 16 Class I events (Pango X lineages unequivocally associated with
a recombination node in the sc2ts ARG) in Table~\ref{tab:pango_x_lineages},
8 have corresponding recombination events (4, at $p=4$).
Of the 5 Type II events
(recombination nodes with more complex assocation with X lineages),
two have close matches, with XBM matching exactly
(including 2 descending BF.3 samples),
and XE/XH is similar (but without including the 2 XH samples in the clade).
Two events in the RIPPLES set correspond to likely false negatives
in sc2ts: XB (Document~\ref{sec:dd_xb})
and XAJ (Document~\ref{sec:dd_xaj}).

Firstly, 408 of these RIPPLES identified recombination events
have one or more parents that are \emph{younger} than the
recombinant sequence based on metadata. As sc2ts enforces strict time ordering,
the parental sequences were simply not available for it to match against
when the recombinant sequence was inserted. Secondly, of the 179
remaining events with matching clades and in which the 
sc2ts node has more than 4 mutations, of the total
1,167 mutations, 452 of these occurred at sites that were 
remapped during post-hoc parsimony. The remaining ``missing''
events are likely attributable to the finer details of the 
differences in tree building approaches between UShER and sc2ts

[FIXME: Discuss  the large effect recombinants inferred by both]

Two further recombinants in the sc2ts ARG that are likely to be 
spurious are important to discuss.
[FIXME - discuss BA.5 (kept as a recombination event but only because we can't 
rewire it post-hoc, owning to subsequent tree building) and the BA.2 event (doesn't
pass QC.) - note this is discussed in more detail in document \ref{sec:spurious-ancestrally-important-recombinants} ]

@@

Table~\ref{tab:ripples_x_lineages} shows a summary of all
recombination events associated with Pango X lineages detected by RIPPLES,
and their relationship with the sc2ts ARG.
Of the 16 Class I events (Pango X lineages unequivocally associated with
a recombination node in the sc2ts ARG) in Table~\ref{tab:pango_x_lineages},
8 have corresponding recombination events (4, at $p=4$).
Of the 5 Type II events
(recombination nodes with more complex assocation with X lineages),
two have close matches.
XBM matches exactly (including 2 descending BF.3 samples),
and XE/XH is similar (but without including the 2 XH samples in the clade).
UShER and sc2ts agree on creating two separate XAM clades, but while sc2ts
makes these both descendants of the Xx event,
RIPPLES creates two independent recombinations.
Similarly, XAE and ??? have exactly matching clades, but RIPPLES
infers two independent recombination events while sc2ts groups them both
under the Xx event. XB is notable as a likely sc2ts false negative that is 
identified by RIPPLES at both $p=3$ and $p=4$. The remaining events 
associated with X lineages output by RIPPLES cannot be easily mapped 
to the sc2ts ARG as one or more descendants is distantly placed.
XAJ is an outlier for sc2ts, being the Pango X lineage identified as a 
non-recombinant with the largest number of mutations 
(Table~\ref{tab:pango_x_lineages}), and a likely false negative.
The remaining 4 events are mixtures of multiple lineages that are 
difficult to reconcile with either the other events in the table 
or the sc2ts ARG.

>>>>

The Delta, BA.1, and BA.2 lineages were 
particularly challenging, however, as the initial emergences of 
these lineages were not densely sampled. In this case,
there is little information for the retrospective sampling grouping mechanism
to work with and it is very difficult to distinguish short-term time-travellers
from true outbreak samples. For these cases, we chose specific
high-quality samples with confident dates as ``seeds`` which are 
unconditionally included in the ARG on the specified date.

% We used more seeds, but in practise it doesn't matter and it's just
% complicating the story.
For Delta
(first detected in December, 2020 in India \cite{ECDCVOCUpdates}),
% More samples were unconditionally added,
% but the earliest are the seeds.
we used two seeds:
one sample (ERR5461562) assigned to B.1.617.1 (Kappa) and
the other (ERR5676810) assigned to B.1.617.2 (Delta).
Kappa is closely related to Delta, and
therefore seeding it helps to reconstruct the origin of Delta.
% See https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian_config.toml
Both of these seeds were sampled in the UK,
the Kappa one in February, 2021 and the Delta one in March, 2021.
% https://github.com/jeromekelleher/sc2ts-paper/issues/267
% The reality is a bit more complicated,
% but this is easy to explain.
These are the earliest Kappa and Delta samples
in the Viridian v04 dataset
that were also used for Pango designation.
% FIXME more Delta details and some references to back up our assertions
% There's some specific info in the Viridian paper on this, I'm sure.
% We should mention long branch splitting here too.]
The resulting subgraph around Delta in the ARG
can be seen in Figure~\ref{fig:subgraph_delta};
for further explanation, see Document~\ref{sec:dd_delta}.

% FIXME: revise this paragraph to give some background info on Omicron,
% and reference the Viridian work]
For Omicron BA.1 and BA.2
(first detected in South Africa or Botswana
in November, 2021 \cite{ECDCVOCUpdates}),
we used SRR17041376 and SRR17461792, respectively.
We chose SRR17041376 (dated November 12, 2021 from South Africa)
as the BA.1 seed
(\url{https://github.com/jeromekelleher/sc2ts-paper/issues/264}),
as it could be well explained as a non-recombinant
given the ARG available on its collection date.
For BA.2,
we chose SRR17461792 (dated November 27, 2021 from South Africa),
which is the earliest BA.2 sample present in the Viridian v04 dataset.
This sample is dated later than SRR17041376 (the BA.1 seed),
allowing time for the BA.1 lineage
to accumulate enough sampled descendants and
to get established in the ARG.
The resulting subgraph around these Omicron lineages in the ARG
can be seen in Figure~\ref{fig:subgraph_omicron}.
For further explanation, see Document~\ref{sec:dd_omicron}.

@@

Using the default ``parsimony improvement'' parameter of $p=3$, UShER+RIPPLES detected a total of 4111 recombination events.
Most of these subtend $<10$ samples (2616 singletons),
although four events subtended more than 100,000 samples.
Analysis of the descending samples for each event shows that
10 of the 44 Pango X lineages present in the dataset
% NOTE: listing them like this for sorting and manipulation purposes
(%
XA, 
XB, 
XC, 
XL, 
XW,
XAC, 
XAE, 
XAJ,
XBB, 
XBM%
)
are associated unequivocally with a single recombination event.
7 lineages 
(%
XE,
XM,
XS,
XY,
XAM,
XBD, 
XBR%
)
are associated with two events. Finally, 
one event subtending 1321 samples suggests that 
XE, 
XH, 
XJ, 
XM, 
XY, 
XAF,
XAL, 
and 
82 BA.2 samples share a recombination event whose parents are BA.2 and BA.1 [FIXME: With what breakpoint and parents?].
We also ran RIPPLES with $p=4$, which reduced the overall number of
events to 1174 (roughly 4-fold) but also approximately halved the number of
events involving X lineages, missing well-characterized events,
such as XBB (Table~\ref{tab:ripples_x_lineages}).

To compare these results with the sc2ts ARG, we first matched clades
by finding the most recent common ancestor (MRCA) of the samples
descending from each RIPPLES event. Overall, clade agreement was
high with [FIXME, get exact numbers] 37XX nodes (90\%) in the sc2ts ARG
where descendants match exactly. Of these nodes, 68\% had no recombinant
ancestry. Of the remainder, 595 could be matched with
one of the 855 recombination events in the sc2ts ARG
(Figure~\ref{fig:ripples_sc2ts_events}).
Table~\ref{tab:ripples_x_lineages} shows a summary of
all recombination events associated with Pango X lineages detected by RIPPLES,
and their relationship with the sc2ts ARG.
Of the 16 Class I events (Pango X lineages unequivocally associated with
a recombination node in the sc2ts ARG) in Table~\ref{tab:pango_x_lineages},
8 are found by RIPPLES (4, at $p=4$).
Of the 5 Type II events
(recombination nodes with more complex assocation with X lineages),
two have close matches in RIPPLES, with XBM matching exactly
(including 2 descending BF.3 samples),
and XE/XH is similar (but without including the 2 XH samples in the clade).
Two events in the RIPPLES set correspond to likely false negatives
in sc2ts: XB (Document~\ref{sec:dd_xb})
and XAJ (Document~\ref{sec:dd_xaj}).

Firstly, 408 of these RIPPLES identified recombination events
have one or more parents that are \emph{younger} than the
recombinant sequence based on metadata. As sc2ts enforces strict time ordering,
the parental sequences were simply not available for it to match against
when the recombinant sequence was inserted. Secondly, of the 179
remaining events with matching clades and in which the 
sc2ts node has more than 4 mutations, of the total
1,167 mutations, 452 of these occurred at sites that were 
remapped during post-hoc parsimony. The remaining ``missing''
events are likely attributable to the finer details of the 
differences in tree building approaches between UShER and sc2ts.

[
FIXME: Discuss  the large effect recombinants inferred by both:
sc2ts identifies two recombination events with \textgreater 10k descendants (specifically, 166,979 and 166,285 descendant samples), both of which we identify as spurious (see \ref{sec:spurious-ancestrally-important-recombinants}). RIPPLES at $p=4$ also identifies two recombination events with \textgreater 10k descendants (531,224 and 339,523 descendant samples); these do not correspond to the same events as in sc2ts. At $p=3$, RIPPLES identifies six recombination events with \textgreater 10k descendant samples, four of which have \textgreater 100k descendant samples.
]

@@

Table~\ref{tab:ripples_x_lineages} shows a summary of all
recombination events associated with Pango X lineages detected by RIPPLES,
and their relationship with the sc2ts ARG.
Of the 16 Class I events (Pango X lineages unequivocally associated with
a recombination node in the sc2ts ARG) in Table~\ref{tab:pango_x_lineages},
8 have corresponding recombination events (4, at $p=4$).
Of the 5 Type II events
(recombination nodes with more complex assocation with X lineages),
two have close matches.
XBM matches exactly (including 2 descending BF.3 samples),
and XE/XH is similar (but without including the 2 XH samples in the clade).
UShER and sc2ts agree on creating two separate XAM clades, but while sc2ts
makes these both descendants of the Xx event,
RIPPLES creates two independent recombinations.
Similarly, XAE and [FIXME] ??? have exactly matching clades, but RIPPLES
infers two independent recombination events while sc2ts groups them both
under the Xx event. XB is notable as a likely sc2ts false negative that is 
identified by RIPPLES at both $p=3$ and $p=4$. The remaining events 
associated with X lineages output by RIPPLES cannot be easily mapped 
to the sc2ts ARG as one or more descendants is distantly placed.
XAJ is an outlier for sc2ts, being the Pango X lineage identified as a 
non-recombinant with the largest number of mutations 
(Table~\ref{tab:pango_x_lineages}), and a likely false negative.
The remaining 4 events are mixtures of multiple lineages that are 
difficult to reconcile with either the other events in the table 
or the sc2ts ARG.
